---
title: "Simulating and Fitting a Spatiotemporal ZINB-GP Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating and Fitting a Spatiotemporal ZINB-GP Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
set.seed(20260729)
```

## The model

`ZINB.GP` models a count in two stages. First, a Bernoulli variable says whether
observation \(j\) is in an at-risk state:

\[
W_j \sim \operatorname{Bernoulli}(\phi_j), \qquad
\operatorname{logit}(\phi_j) =
X_j\alpha + V_{s,j}a + V_{t,j}b.
\]

If \(W_j=0\), the count is a structural zero. If \(W_j=1\), the count follows a
negative-binomial distribution:

\[
Y_j \mid W_j=1 \sim \operatorname{NB}(r,\psi_j), \qquad
\operatorname{logit}(\psi_j) =
X_j\beta + V_{s,j}c + V_{t,j}d.
\]

With the package's parameterization,
\(\operatorname{E}(Y_j\mid W_j=1)=r\exp(\eta_{2j})\). Thus `Alpha`
describes the at-risk probability and `Beta` describes conditional count
intensity. The marginal expected count is the product of those two quantities.

The vectors \(a,c\) are spatial effects and \(b,d\) are temporal effects. Each
has a noisy squared-exponential GP prior

\[
\sigma^2\{\kappa K_\ell + (1-\kappa)I\}.
\]

Here, \(\sigma^2\) is marginal variance, \(\ell\) is a length scale, and
\(\kappa\) is the fraction of variance attributed to structured dependence.

## Build a correctly paired space-time design

The rows of `Vs`, `Vt`, `X`, and `y` must refer to the same observations. The
helper below samples a number of replicates for every spatial-temporal cell and
then repeats the paired cell indices. This construction gives every row of
`Vs` and `Vt` a shared space-time identity and keeps both matrices aligned with
the corresponding rows of `X` and `y`.

```{r design}
library(Matrix)
library(mvtnorm)
library(ZINB.GP)

make_design <- function(n_space, n_time, mean_replicates) {
  cell_n <- matrix(
    rpois(n_space * n_time, mean_replicates),
    nrow = n_space
  )
  cells <- expand.grid(
    spatial = seq_len(n_space),
    temporal = seq_len(n_time)
  )

  spatial_id <- rep(cells$spatial, times = as.vector(cell_n))
  temporal_id <- rep(cells$temporal, times = as.vector(cell_n))
  n <- length(spatial_id)

  Vs_full <- as.matrix(sparseMatrix(
    i = seq_len(n), j = spatial_id, x = 1,
    dims = c(n, n_space)
  ))
  Vt_full <- as.matrix(sparseMatrix(
    i = seq_len(n), j = temporal_id, x = 1,
    dims = c(n, n_time)
  ))

  list(Vs_full = Vs_full, Vt_full = Vt_full, cell_n = cell_n)
}

n_space <- 12
n_time <- 8
design <- make_design(n_space, n_time, mean_replicates = 3)
N <- nrow(design$Vs_full)
N
```

## Baselines and intercepts

An intercept together with every column of both one-hot indicator matrices is
rank deficient. `ZINB_GP_orig()`, and the full-GP route through `ZINB_GP()`, use
the first spatial level and first temporal level as baselines:

- `X` contains an explicit intercept;
- the first columns of `Vs_full` and `Vt_full` are removed;
- the first rows and columns of the distance matrices are removed internally;
- the remaining GP effects are sampled directly from the corresponding reduced
  covariance matrices.

The simulation therefore draws the \(S-1\) spatial and \(T-1\) temporal effects
directly from these reduced GP covariance matrices. The fixed intercept is the
linear predictor for an observation at both baseline levels, and each retained
effect is the additive departure associated with its nonbaseline level. This
parameterization gives a full-rank design and matches the covariance model used
for estimation.

```{r distances}
coords <- cbind(runif(n_space), runif(n_space)) * 1000
time_coord <- matrix(0:(n_time - 1) * 50, ncol = 1)

Ds <- as.matrix(dist(coords))
Dt <- as.matrix(dist(time_coord))

Vs <- design$Vs_full[, -1, drop = FALSE]
Vt <- design$Vt_full[, -1, drop = FALSE]

stopifnot(
  ncol(Vs) + 1 == nrow(Ds),
  ncol(Vt) + 1 == nrow(Dt),
  nrow(Vs) == nrow(Vt)
)
```

The coordinate units are deliberately large enough for the package's
pre-MCMC kernel-conditioning screen. The generating length scales below are
expressed in those same units, so this numerical scaling does not change the
intended correlation pattern.

## Draw latent effects and counts

The package squares the supplied distances before applying its default kernel.
We do the same when generating the latent effects. The reduced covariance
matrices exactly match the identified model fitted by `ZINB_GP_orig()`.

```{r simulate}
noisy_covariance <- function(distance, length_scale, sigma, kappa) {
  correlation <- exp(-(distance^2) / length_scale^2)
  sigma^2 * (kappa * correlation + (1 - kappa) * diag(nrow(distance)))
}

spatial_distance <- Ds[-1, -1, drop = FALSE]
temporal_distance <- Dt[-1, -1, drop = FALSE]

a <- drop(rmvnorm(1, sigma = noisy_covariance(
  spatial_distance, length_scale = 350, sigma = 1, kappa = 0.5
)))
c <- drop(rmvnorm(1, sigma = noisy_covariance(
  spatial_distance, length_scale = 250, sigma = 1, kappa = 0.5
)))
b <- drop(rmvnorm(1, sigma = noisy_covariance(
  temporal_distance, length_scale = 100, sigma = 0.5, kappa = 0.2
)))
d <- drop(rmvnorm(1, sigma = noisy_covariance(
  temporal_distance, length_scale = 150, sigma = 0.5, kappa = 0.2
)))

x <- rnorm(N)
X <- cbind("(Intercept)" = 1, x = x)
alpha <- c(-0.25, 0.25)
beta <- c(0.50, -0.25)
r <- 1

eta_at_risk <- drop(X %*% alpha + Vs %*% a + Vt %*% b)
p_at_risk <- plogis(eta_at_risk)
at_risk <- rbinom(N, size = 1, prob = p_at_risk)

eta_count <- drop(X %*% beta + Vs %*% c + Vt %*% d)
mu_count <- r * exp(eta_count)
y <- integer(N)
y[at_risk == 1] <- rnbinom(
  sum(at_risk == 1),
  size = r,
  mu = mu_count[at_risk == 1]
)

c(observations = N, zeros = sum(y == 0), positive = sum(y > 0))
```

The heatmap below is a useful first check. It shows cell totals rather than
individual replicates, making both zero inflation and clusters of large counts
visible on the modeled support.

```{r simulated-heatmap}
cell_id <- max.col(design$Vs_full) +
  n_space * (max.col(design$Vt_full) - 1)
cell_sum <- tapply(y, cell_id, sum)
cell_total <- numeric(n_space * n_time)
cell_total[as.integer(names(cell_sum))] <- cell_sum
cell_total <- matrix(cell_total, nrow = n_space)

image(
  x = seq_len(n_time),
  y = seq_len(n_space),
  z = t(log1p(cell_total)),
  xlab = "Time",
  ylab = "Spatial location",
  main = "Simulated cell totals: log(1 + count)",
  col = hcl.colors(20, "YlOrRd", rev = TRUE)
)
```

## Fit the full model

The following call uses GPs in both components. It is not evaluated while the
vignette is built because a useful MCMC run is intentionally much longer than a
CRAN vignette should take.

```{r fit, eval=FALSE}
fit <- ZINB_GP(
  X = X,
  y = y,
  coords = coords,
  Vs = Vs,
  Vt = Vt,
  Ds = Ds,
  Dt = Dt,
  nsim = 20000,
  burn = 5000,
  thin = 5,
  save_ypred = TRUE,
  print_progress = TRUE,
  use_count_gp = TRUE,
  use_inflation_gp = TRUE
)
```

The rows of `coords` identify the full set of spatial levels, including the
baseline. `Ds` and `Dt` control the dense GP covariances. Distance and length
scale share the same units: multiplying a distance matrix by a constant requires
multiplying its length scale by that constant to preserve the same covariance.

The entry point also supports simpler models. For example, this call keeps the
spatial and temporal GPs only in the count component:

```{r count-only, eval=FALSE}
fit_count_gp <- ZINB_GP(
  X = X, y = y, coords = coords,
  Vs = Vs, Vt = Vt, Ds = Ds, Dt = Dt,
  nsim = 20000, burn = 5000, thin = 5,
  use_count_gp = TRUE,
  use_inflation_gp = FALSE
)
```

The component flags encode which residual dependence the model represents. A
count GP explains residual variation in conditional intensity; an inflation GP
explains residual variation in whether an observation is at risk. The selected
combination should follow the scientific role assigned to each process.

## Interpret posterior draws

Each row of `fit$Alpha` and `fit$Beta` is a retained posterior draw. The random
effects are in `A` and `B` for the at-risk component and `C` and `D` for the
count component. `R` contains dispersion draws, and `Noise1s`, `Noise1t`,
`Noise2s`, and `Noise2t` are the structured-variance fractions \(\kappa\).

For a fitted model, equal-tailed intervals can be computed directly:

```{r intervals, eval=FALSE}
apply(fit$Alpha, 2, quantile, probs = c(0.025, 0.5, 0.975))
apply(fit$Beta, 2, quantile, probs = c(0.025, 0.5, 0.975))
```

Credible intervals summarize posterior uncertainty; they do not establish that
the Markov chain mixed well. The Oregon case-study vignette shows how to wrap
these matrices with standard MCMC packages and calculate effective sample size.
